Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex
Update paper.tex

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and GPUs.

@@

To demonstrate the ease and efficiency with this data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. In the first we re-implement the af-dist calculation
used throughout as an example of a non-trivial whole-matrix operation
on GPU. In the second, we demonstrate some emerging technology
in which Zarr data can be decompressed directly in GPU memory.
In both cases we demonstrate on chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 variants.
Benchmarks in this section were performed on a
machine with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

We first illustrate the ease with which computations can be
be peformed on the GPU, with little or no preprocessing of the
data. To do this we reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the core operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which
is also very similar to the Numba compiled function used in other
benchmarks. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82MiB), decompress each to a local
NumPy array and then copying the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with an
overall data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer
time to 138ms and processing time to 6ms.

Chunk decompression comprises a significant fraction of processing
time [FIXME crossrefs?] and host-to-GPU data transfer is
often a bottleneck. It is therefore useful in some applications
to directly decompress Zarr chunks into GPU memory.
KvikIO~\cite{kvikio} is a Python package that provides access to optimised
GPU IO routines, and providing support for Zarr along with
access to several GPU optimised compression codecs.
To explore the performance of these codecs we recompressed
the \texttt{call\_AD} (allele depth) array
(96514, 2504, 7) [int16 more details]. In all cases it took around 1s
to compress and store this 3.15GiB array. [Details about compression codecs?]
We then measured the time to decompress these arrays and
copy back to host memory. This required about 0.6s with the
fastest codecs. For reference, using the Zstandard codec and the
standard CPU Zarr code (which does some limited multithreading for decompression
by default) required about 7.6s. [Some discussion of how to interpret
these benchmarks. We're just trying to illustrate the powerful toolchain
 that is immediately to hand when data is stored in Zarr format.]

>>>>

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.

@@

To demonstrate the ease and efficiency with this data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. In the first we re-implement the af-dist calculation
used throughout as an example of a non-trivial whole-matrix operation
on GPU. In the second, we demonstrate some emerging technology
in which Zarr data can be decompressed directly in GPU memory.
In both cases we demonstrate on chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 variants.
Experiments in this section were performed on a
machine with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

We first illustrate the ease with which computations can be
be peformed on the GPU, with little or no preprocessing of the
data. To do this we reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the core operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which
is also very similar to the Numba compiled function used in other
experiments. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82MiB), decompress each to a local
NumPy array and then copying the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with an
overall data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer
time to 138ms and processing time to 6ms.

Chunk decompression comprises a significant fraction of processing
time [FIXME crossrefs?] and host-to-GPU data transfer is
often a bottleneck. It is therefore useful in some applications
to directly decompress Zarr chunks into GPU memory.
KvikIO~\cite{kvikio} is a Python package that provides access to optimised
GPU IO routines, and providing support for Zarr along with
access to several GPU optimised compression codecs.
To explore the performance of these codecs we recompressed
the \texttt{call\_AD} (allele depth) array
(96514, 2504, 7) [int16 more details]. In all cases it took around 1s
to compress and store this 3.15GiB array. [Details about compression codecs?]
We then measured the time to decompress these arrays and
copy back to host memory. This required about 0.6s with the
fastest codecs. For reference, using the Zstandard codec and the
standard CPU Zarr code (which does some limited multithreading for decompression
by default) required about 7.6s. [Some discussion of how to interpret
these experiments. We're just trying to illustrate the powerful toolchain
 that is immediately to hand when data is stored in Zarr format.]
